# Recap of class 3

- Recap of class 3 topics
- Class 3 homework review

At this point you should have enough knowledge to import, explore, visualize, and wrangle data in R! You should also have a good understanding of all the fundamentals you need to branch out and start learning how to do more specialized tasks on your own. Those are all the tough parts - from here on out, classes can focus on more practical applications of the tool.

# Data Analysis Fundamentals

- The most important part in answering a question is: Clearly defining the question!
- Data comes second - identify and collect it
- Data cleaning is commonly overlooked but absolutely critical for valid conclusions
- Once you have a question and a relevant, clean dataset then run your analysis by creating models, testing hypotheses, building visualizations, etc.
- If you come up with predictions, attempt to validate them
- Finally, create your output and document your work - make it as reproducible as possible/practical

# Practical exercises

Note: From now on, it's assumed you understand how to install packages if you do not already have them installed. Also you should be able to find and review relevant documentation.

In this practice exercise, will answer the questions or complete tasks posed below using datasets from the nycflights13 library:
Q1: What was the fastest NYC flight in 2013?
Q2: What airlines are most delayed on average?
Q3: Create a chart that visually describes delays by airline
Q4: Do delays correlate with greater speed in-flight?
Q5: Does weather significantly affect flight delay severity?

Load required packages

```{r}
library(tidyverse)
library(nycflights13)
```

Explore the data

```{r}
?flights
df <- flights
summary(df)
str(df)
```

Q1: What was the fastest NYC flight in 2013?

```{r}
df %>%
  mutate(speed = distance/air_time) %>%
  filter(speed == max(speed, na.rm = TRUE)) %>%
  View()
```

Q2: What airlines are most delayed on average?

```{r}
df %>%
  mutate(total_delay = dep_delay + arr_delay) %>%
  group_by(carrier) %>%
  summarise(Average_Delay = mean(total_delay, na.rm = TRUE)) %>%
  left_join(airlines) %>%
  select(Airline = name,
         Airline_Code = carrier,
         Average_Delay) %>%
  arrange(desc(Average_Delay))
```

Q3: Create a chart that visually describes delays by airline

```{r}
carrier_dep_delay <- df %>%
  mutate(total_delay = dep_delay + arr_delay) %>%
  filter(!is.na(total_delay))

ggplot(data = carrier_dep_delay, aes(x = total_delay, fill = carrier)) + 
  geom_density(alpha = .3)

# For easier viewing, remove all departures >=100min
carrier_dep_delay %>%
  filter(total_delay < 100) %>%
  ggplot(data = ., aes(x = total_delay, fill = carrier)) + 
    geom_density(alpha = .3)
```

Q4: Do delays correlate with greater speed in-flight?

```{r}
df <- df %>%
  mutate(total_delay = dep_delay + arr_delay) %>%
  mutate(speed = distance/air_time)

cor_x <- df %>%
  filter(!is.na(total_delay)) %>%
  select(total_delay)

cor_y <- df %>%
  filter(!is.na(speed)) %>%
  select(speed)

# Do a correlation for the two variables
cor(cor_x, cor_y)

# Create a visualization
ggplot(data = df, aes(x = speed, y = total_delay)) +
  geom_point() +
  geom_smooth()
```

Q5:Does weather significantly affect flight delay severity?

```{r}
# TBD
```

# Homework

For this class, complete the following homework assignments:

- Finish answering any open questions from this problem set; otherwise, ask more questions that could be answered with the nycflights13 dataset and practice answering them
- Read Tidy Data and A Layered Grammar of Graphics, both by Hadley Wickham
- Now that your data is imported and cleaned up, begin working on the script that will answer your research question; you should have everything you need to extract conclusions related to your project!
