# Recap of class 5

- Recap of class 5 topics
- Class 5 homework review

# Why create functions?

Functions are the workhorses of the R programming language. They help speed up or automate workflows by simplying repetitive tasks. They are easier to use than typical R scripts that perform a task, because a well-defined function will have some documentation on what inputs it accepts, what output values the user can expect, and what it does exactly. They can also be called with simple, descriptive names and include error checking. Finally, functions can be nested (for example with mutate) to become more powerful than when they are used independently. The fact that they use arguments to contorl inputs means that a well-defined function can be more flexible than a R script since possible alternate uses for it are pre-built.

# Function example: spcCharts

This function is used to automate an otherwise lengthy workflow that accepts a CSV input, mutates it extensively, calculates statistical process control charts with another function, and finally contructs a ggplot2 to chart.

```{r}
# Function accepts a data file, start and end dates, and grouping
# Returns SPC chart
# Function expects certain standards on data file:
## First column contains dates in the format "Mmm, YYYY"
## Second column contains a count of events/issues/deficiencies
## Third column contains the number of relevant operations
## There should be no skipped months within dataset
## Oldest date first (top), newest date last (bottom)
# Data is collected at month end; mid-month date ranges will exclude the last month

spcChart <- function(filename = NA,
                     title = NA,
                     group = c("monthly", "quarterly"),
                     enddate = floor_date(today(), "month") - 1, # Last day of previous month
                     startdate = enddate - months(24) + 1) { # Go 24 months back

  # Required packages without exposing messages
  suppressMessages(require(lubridate))
  suppressMessages(require(tidyverse))
  suppressMessages(require(qcc))

  # Read input files
  df <- read.csv(filename, stringsAsFactors = FALSE)

  # Check for legitimate arguments
  if(!(is.data.frame(df)) | ncol(df) != 3) {
    stop("Data file not recognized as a data.frame with three columns")
  }
  if(!(group[1] %in% c("monthly", "quarterly"))) {
    stop("Grouping type not recognized")
  }
  if(!is.Date(startdate) | !(is.Date(enddate))) {
    stop("Start and end dates not recognized")
  }

  # Clean up input data file
  # Fix column names based on position
  df <- df %>%
    select(Date_Label_Month = 1,
           Events = 2,
           Operations = 3)

  # Convert date labels into R Date format, necessary for grouping and filtering
  dates_list <- df %>%
    select(Date_Label_Month) %>%
    unlist() %>%
    strsplit(",")

  months_vector <- dates_list %>%
    sapply("[", 1) %>%
    unname()

  years_vector <- dates_list %>%
    sapply("[", 2) %>%
    unname() %>%
    as.integer()

  dates_vector <- paste(years_vector, months_vector, "15", sep = "-") %>%
    as.Date(format = "%Y-%b-%d") %>%
    ceiling_date("month") - 1

  df <- df %>%
    mutate(Date_Collected = dates_vector) %>%
    mutate(Quarter_int = quarter(Date_Collected)) %>%
    mutate(Year_int = year(Date_Collected)) %>%
    mutate(Date_Label_Quarter = paste(paste("Q", Quarter_int, sep = ""),
                                      Year_int, sep = ", ")) %>%
    select(-Quarter_int, -Year_int)

  # Apply filter
  df <- df %>%
    filter(Date_Collected >= startdate, Date_Collected <= enddate)

  # Apply grouping
  if(group[1] == "monthly") {
    # Note that data is already grouped by month from input
    # No additional grouping operation is required here
    df_grouped <- df %>%
      select(Date_Label = Date_Label_Month,
             Events = Events,
             Operations = Operations)
  }
  if(group[1] == "quarterly") {
    df_grouped <- df %>%
      group_by(Date_Label_Quarter) %>%
      summarise(Events = sum(Events), Operations = sum(Operations)) %>%
      ungroup() %>%
      select(Date_Label = Date_Label_Quarter,
             Events = Events,
             Operations = Operations)

    # Must re-order data.frame after grouping to preserve time series order
    dates_list_ordering <- df_grouped %>%
      select(Date_Label) %>%
      unlist() %>%
      strsplit(", ")

    quarters_vector_ordering <- dates_list_ordering %>%
      sapply("[", 1) %>%
      unname()

    years_vector_ordering <- dates_list_ordering %>%
      sapply("[", 2) %>%
      unname() %>%
      as.integer()

    df_grouped <- df_grouped %>%
      mutate(Year_int = years_vector_ordering) %>%
      mutate(Quarter_int = quarters_vector_ordering) %>%
      arrange(Year_int, Quarter_int) %>%
      select(-Year_int, -Quarter_int)
  }

  # Create quality control chart
  # Collect variables for QCC calculation
  observations <- df_grouped %>%
    select(Events) %>%
    unlist() %>%
    unname()

  labels <- df_grouped %>%
    select(Date_Label) %>%
    unlist() %>%
    unname()

  sizes <- df_grouped %>%
    select(Operations) %>%
    unlist() %>%
    unname()

  # Calculate QCC values
  qcc_calc <- qcc(data = observations, type = "np", sizes = sizes, plot = FALSE)

  qcc_df <- data.frame(labels = labels,
                           center = qcc_calc$center,
                           values = qcc_calc$statistics,
                           LCL = qcc_calc$limits[,1],
                           UCL = qcc_calc$limits[,2],
                           beyond_limits = 0,
                           stringsAsFactors = FALSE)

  # Convert labels into factor so ggplot will know to order it non-alphabetically
  qcc_df$labels <- factor(qcc_df$labels,
                              levels = unique(qcc_df$labels))

  # Update data.frame with rows where runs rules were exceeded
  runs_index <- unname(unlist(qcc_calc$violations[2]))
  qcc_df[runs_index, "beyond_limits"] <- 2

  # Update data.frame with rows where limit was exceeded
  beyond_limits_index <- unname(unlist(qcc_calc$violations[1]))
  qcc_df[beyond_limits_index, "beyond_limits"] <- 1

  # Create ggplot
  plot <- ggplot(qcc_df) +
    labs(title = title, x = "Dates", y = "Count") +
    theme(axis.text = element_text(size = 8),
          axis.title = element_text(size = 14),
          title = element_text(size = 18)) +
    geom_step(aes(x = labels, y = LCL), group = 1, linetype = 2, color = "red") +
    geom_step(aes(x = labels, y = UCL), group = 1, linetype = 2, color = "red") +
    geom_step(aes(x = labels, y = center), group = 1, linetype = 2, color = "blue") +
    geom_line(aes(x = labels, y = values), group = 1) +
    geom_point(aes(x = labels, y = values, color = factor(beyond_limits)), size = 4) +
    scale_colour_manual(values = c("0" = "black", "1" = "red", "2" = "yellow"), guide = FALSE)

  # Plot
  return(plot)
  }
```

The spcChart_all uses the spcChart function on all existing CSV files in working directory and saves them as PNG files.

```{r}
# Uses spcChart function on all existing CSV files in working directory
# Saves PNG files

spcChart_all <- function() {
  
  # Load required libraries
  require(safetydata)
  
  # Get list of files that will be used for chart creation
  files <- list.files(pattern = ".csv")
  
  for(i in 1:length(files)) {
    title <- unlist(strsplit(files[i], ".csv"))
    
    # Generate chart
    spc_Plot <- spcChart(filename = files[i],
                         title = title)
    
    # Save image file
    ggsave(filename = paste(title, "png", sep = "."),
           plot = spc_Plot,
           device = "png",
           path = getwd(),
           height = 7,
           width = 14,
           units = "in")
  }
}
```

# Practical exercises

Basic function components
- Arguments: order & defaults are important
- Input error checking
- Return a value at the end

The example below is a simple function template

```{r}
fun <- function(arg1, arg2 = "default") {
	# function goes here
	# loops
	if(exists(value)) {
		# blah blah blah
	  # this can be used for error checking and stopping
	stop("Error!")}
	return()
}
```

The example below is a simple function that includes error checking

```{r}
# The purpose of the addTwo function is to add two numbers
addTwo <- function(x, y) {
  # Check for correct data types and lenghts
  if(!is.numeric(x) | !is.numeric(y)) {
    stop("Both x and y must be numeric vectors.")
  }
  if(length(x) != 1 | length(y) != 1) {
    stop("Both x and y must be length 1 vectors.")
  }
  
  # Calculate added value
  z <- x + y
  
  # Return added value
  return(z)
}
```

The example below is a simple function that includes error checking and looping

```{r}
# The purpose of the nameAndPaste function is to grab all files in your
# Working directory with a given extension, and return all names
# Pasted together as one line
nameAndPaste <- function(extension = ".csv") {
  # Error checking
  if(!is.character(extension)) {
    stop("Extension must be expressed as a character")
  }
  
  # Get all file names
  files <- list.files(pattern = extension)
  
  # Create a blank pasted_file_names object
  pasted_file_names <- NULL
  
  # Loop through file names pasting them
  for(i in 1:length(files)) {
    current_file_name <- unlist(strsplit(files[i], extension))
    pasted_file_names <- paste(pasted_file_names, current_file_name, sep = "/")
  }
  
  # Returns a vector of file names
  return(pasted_file_names)
}
```

# Homework

- Create a function that automates a repetitive part of your workflow, for example takes an input and builds a graph that you've had to build by hand in the past
- Complete and email me your capstone analysis; prepare a 5/10-minute presentation on it for the final class
