# Recap of class 3

- Recap of class 3 topics
- Class 3 homework review

# ggplot2

The grammar of graphics. Uses a small number of common "verbs" and very standardized syntax to enable the programmer to describe any chart they can imagine.

data + coordinates [mapping] + geometric shapes = plot

# Canonical plots

- One continuous variable: Histogram (density)
- One discrete variable: Barplot
- Two continuous variables: Scatterplot
- One discrete and one continuous variables: Boxplot
- Two discrete variables: Heat map
- Time series: Line

```{r}
# Load libraries
library(tidyverse)
library(yarrr)
library(lubridate)
library(GGally)
library(plotly)
```

```{r}
# One variable = geom_histogram (discrete or cont binned), geom_density (continuous)
ggplot(data = df, mapping = aes(x = Type1)) + geom_bar() # Discrete
ggplot(data = df, mapping = aes(x = speed)) + geom_histogram() # Continuous
ggplot(data = df, mapping = aes(x = speed)) + geom_density() # Continuous

# Cont vs cont = geom_point, geom_smooth
ggplot(data = df, mapping = aes(x = attack, y = defense)) + geom_point()
ggplot(data = df, mapping = aes(x = attack, y = defense)) + geom_point() + geom_smooth(method = "lm", se = FALSE)

# Discrete vs cont = geom_boxplot
ggplot(data = df, mapping = aes(x = Type1, y = hp)) + geom_boxplot()

# Time series = geom_line - Note: Someone correctly pointed out that this plot is in the incorrect order [not fixed]
ggplot(data = df, mapping = aes(x = paste(year(date_caught), month(date_caught), sep = "-"), group = 1)) +
  geom_line(stat = "count")
```

# Homework

- Review the [data visualization cheat sheet](https://www.rstudio.com/resources/cheatsheets/)
- Build visualizations for your data analysis
- Read Tidy Data and A Layered Grammar of Graphics, both by Hadley Wickham
