# Recap of class 1

- Recap of class 1 topics
- Class 1 homework review

# Functions

Functions are the workhorses of programming languages. They take inputs, called arguments, and yield outputs after having performed some action on it.

- Arguments
- Order of arguments
- Default arguments
- Required arguments
- Ellipsis

# Operators

Operators are used to evaluate a logical statement. Using an operator is similar to using a function, except that the syntax is different. Using an operator in R yields a logical vector of evaluations.

- Equal to: ==
- Not equal to: !=
- Greater than, less than: >, <
- Greater than or equal to, less than or equal to: =>, <=
- And, or: &, |
- In list: %in%
- Opposite: !

# Tidyverse

Group of packages created and maintained by Hadley Wickam that provide an easy way to import (readr), wrangle (dplyr/tidyr), organize (tibble), and visualize (ggplot2) your data using the "grammar of data science." The packages work seamlessly together with each other and there is a vibrant community of contributors who keep updates coming frequently.

In my opinion, it superior to base R and makes interacting with data using R far easier and more enjoyable. Everything that can be done in the tidyverse can be done in base R with analogous functions.

# Practical exercises

Load packages and doing essential data exploration

```{r}
library(tidyverse)
library(yarrr)

# Copy dataset to Global Environment
df <- pirates

# Basic exploration
str(df)
summary(df)

# What do my columns mean?
# ?pirates
```

Subsetting rows and columns (selecting and filtering)

```{r}
# Selecting columns and rows
# ?select

# Create an additional copy of the dataset so you don't have to constantly re-load it
df2 <- df
df2 <- select(df2, id, height, weight) # Select columns by name

df2 <- df
df2 <- select(df2, 1, 4, 5) # Select columns by index

# ?filter

df2 <- df
df2 <- filter(df2, sex == "male") # Equal to

df2 <- df
df2 <- filter(df2, sex != "female") # Not equal to

df2 <- df
df2 <- filter(df2, sword.time < 5) # Less than

# The unique or table functions are useful to determine whether to use ==, or !=, or %in%
unique(select(df2, sex))
table(select(df2, sex))
```

Combining select and filter

```{r}
# Filter by > 1 col
df2 <- df
df2 <- filter(df2, sex == "male", sword.time < 5)

# Nested functions
df2 <- df
df2 <- select(filter(df2, sex == "female", eyepatch == 1), id, age, headband)
```

Creating calculated columns

```{r}
# Creating new columns
# ?mutate

df2 <- df
df2 <- mutate(df2, height_in = height / 2.54)

# Drop height column
df2 <- df
df2 <- select(df2, -height)
```

Pipes

```{r}
# This works, but is resource intensive and annoying to read/write
# This is also difficult to alter (for example if I wanted to alter df2 to df)
df2 <- df
df2 <- mutate(df2, height_in = height / 2.54)
df2 <- filter(df2, sex == "female", eyepatch == 1)
df2 <- select(df2, id, age, headband)
df2 <- mutate(df2, new_id = id * 10)
df2 <- select(df2, -id)

# This is impossible to read and troubleshoot
df2 <- df
df2 <- select(mutate(select(filter(mutate(df2,
                                          height_in = height / 2.54),
                                   sex == "female", eyepatch == 1), 
                            id, age, headband),
                     new_id = id * 10),
              -id)

# That is why we use the %>% (piping) operator
df2 <- df
df2 <- df2 %>%
  mutate(height_in = height / 2.54) %>%
  filter(sex == "female", eyepatch == 1) %>%
  select(id, age, headband) %>%
  mutate(new_id = id * 10) %>%
  select(-id)
```

Loading data: We'll use CSV here as a very common file extension for data storage. JSON, XML, SQL, etc. can all be used with different R packages. RData is a good R-specific data storage extension for fast reading/writing, even in large files.

```{r}
x <- read_csv("~/Documents/R_Working_Directory/student_data.csv")
```

# Homework

For this class, complete the following homework assignments:

- Review the [data import and data transformation cheat sheets](https://www.rstudio.com/resources/cheatsheets/)
- Complete Swirl R Programming Basic Building Blocks, lessons 5-9
- Practice loading data from your selected data analysis and selecting/filtering/mutating on it with pipes
